---
title: "Choice over Time: Evidence that Contradicts the Standard Model"
subtitle: "EC404; Spring 2024"
author: "Prof. Ben Bushong"
date: "Last updated `r format(Sys.Date(), '%B %d, %Y')`"
output:
  xaringan::moon_reader:
    lib_dir: libs
    yolo: false
    css: [default, metropolis, "EC404.css"]
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
    incremental: true
name: "Cover"


---
```{r setup, include=FALSE}

options(htmltools.dir.version = FALSE)
options("getSymbols.warning4.0"=FALSE)


library(here) # creates absolute paths, but those won't work in html
library(knitr)
library(kableExtra)
library(tidyverse)
library(wooldridge)
require(lubridate)
require(scales)
require(broom)
require(visualize)

require(wbstats)
require(lmtest)
require(sandwich)
require(car)
require(quantmod)
require(patchwork)



# https://yihui.org/knitr/options/
opts_chunk$set(
  fig.align="center",
  #dpi=300,
  fig.path='figs/', # where figs are rendered
  cache=F,
  echo=F,
  message = F,
  warning = F
  )

```

layout: true

<div class="msu-header"></div>
<div style = "position:fixed; visibility: hidden">
$$\require{color}\definecolor{yellow}{rgb}{1, 0.8, 0.16078431372549}$$
$$\require{color}\definecolor{orange}{rgb}{0.96078431372549, 0.525490196078431, 0.203921568627451}$$
$$\require{color}\definecolor{MSUgreen}{rgb}{0.0784313725490196, 0.52156862745098, 0.231372549019608}$$
</div>

<script type="text/x-mathjax-config">
MathJax.Hub.Config({
  TeX: {
    Macros: {
      yellow: ["{\\color{yellow}{#1}}", 1],
      orange: ["{\\color{orange}{#1}}", 1],
      MSUgreen: ["{\\color{MSUgreen}{#1}}", 1]
    },
    loader: {load: ['[tex]/color']},
    tex: {packages: {'[+]': ['color']}}
  }
});
</script>

<style>
.yellow {color: #FFCC29;}
.orange {color: #F58634;}
.MSUGreen {color: #14853B;}
</style>


```{r flair_color, echo=FALSE}
library(flair)
yellow <- "#FFCC29"
orange <- "#F58634"
MSUGreen <- "#14853B"
```


```{r, eval=FALSE}

#BELOW IS WHERE YOU ADD CONTENT. KEEP ABOVE CONSTANT EXCEPT TITLE.

```

---
class: MSU
# General Version of Discounted Utility

Recall the general version of the discounted-utility model:


$$U^{t}=\sum_{x=0}^{T-t}D(x)\text{ }u_{t+x}.$$

  - $U^{t}$ is intertemporal utility from perspective of period $t$. 
  
  - $u_{\tau }$ is instantaneous utility in period $\tau$ ("well-being" in period $t$). 
  
	- $x$ is the delay before receiving some utility. 
	
	- $D(x)$ is a discount function that specifies the amount of discounting associated with delay $x$. 

Again, in principle, we could have any discount function. 

Exponential discounting **assumes** $D(x)=\delta ^{x}$.

---
class: MSU
# How to Measure Discount Functions?

Typical procedure elicits indifference points of the form:


$(A$ at date $x)$ $\sim$ $(B$ at date $x+y)$ 

--

To interpret, typically make four assumptions (aka **The Usual Assumptions**) 


1. People obey discounted utility model. 

2. People treat these amounts as "bursts" of consumption. 

3. Utility is linear in the amount. 

4. Normalize $D(0)=1$. 

---
class: MSU
# How to Measure Discount Functions?

An implication of the usual assumptions:

$$(A\text{ now})\text{ }\sim \text{ }(B\text{ at date }x)$$

--

$$\Longleftrightarrow D(0)u(A)=D(x)u(B)$$
--

$$A=D(x)B$$ 
--

$$D(x)=\frac{A}{B}$$

---
class: MSU 
# How to Measure Discount Functions?

More generally, the usual assumptions imply:

$$(A\text{ at date }x)\text{ }\sim \text{ }(B\text{ at date }x+y)$$

$$ \Longleftrightarrow \qquad D(x)u(A)=D(x+y)u(B)$$ 
--

$$D(x)A=D(x+y)B$$

--

$$ \frac{D(x+y)}{D(x)}=\frac{A}{B}$$

---
class: MSU 
# How to Measure Discount Functions?

An alternative procedure elicits WTP now for something to be received later. 

E.g., if WTP up to $A$ now to obtain $B$ at date $x$, this implies: 

$(-A$ now & $+B$ at date $x)$ $\sim$ $($ no changes $)$ 

--

Applying the usual assumptions here yields:

$$D(0)u(-A)+D(x)u(B)=0$$ 

--

$$D(x)B=A$$

$$D(x)=\frac{A}{B}$$

---
class: MSU 
# Some Evidence from Thaler (1981)

Using a (hypothetical) matching technique, Thaler found that people were indifferent between: 

| Option | Comparison                              |
|--------|-----------------------------------------|
| a.     | ($15 now) ~ ($30 in 3 months)           |
| b.     | ($15 now) ~ ($60 in 1 year)             |
| c.     | ($15 now) ~ ($100 in 3 years)           |

--

Implications under the usual assumptions: 

a. $D(3$ months $)=\frac{15}{30}=0.50$. 

b. $D(1$ year $)=\frac{15}{60}=0.25$. 

c. $D(3$ years $)=\frac{15}{100}=0.15$.

---
class: MSU
# Interpreting Thaler's Evidence

We can convert each $D(x)$ into an implicit yearly discount rate: 

**Definition:** The *average yearly discount rate* applied to delay $x$ (where $x$ is specified in years) is the $\rho$ such that 

$$e^{-\rho x}=D(x)\qquad \text{or} \qquad\rho =\frac{1}{x}(-\ln D(x)).$$
--

Applying this definition:

a. $D(3$ months $)=0.50$ $\Longrightarrow$ 277% yearly discounting. 

b. $D(1$ year $)=0.25$ $\Longrightarrow$ 139% yearly discounting. 

c. $D(3$ years $)=0.15$ $\Longrightarrow$ 63% yearly discounting.


---
class: MSU
# More Evidence from Thaler (1981)

Thaler (1981) also found that people were indifferent between: 

| Option | Comparison                              |
|--------|-----------------------------------------|
| e.     | ($250 now) ~ ($300 in 3 months)         |
| f.     | ($250 now) ~ ($350 in 1 year)           |
| g.     | ($250 now) ~ ($500 in 3 years)          |

--

Implications under the usual assumptions: 

e. $D(3$ months $)=\frac{250}{300}=0.83$ $\Longrightarrow$ 73% yearly discounting. 

f. $D(1$ year $)=\frac{250}{350}=0.71$ $\Longrightarrow$ 34% yearly discounting. 

g. $D(3$ years $)=\frac{250}{500}=0.50$ $\Longrightarrow$ 23% yearly discounting. 


---
class: MSU 
# Conclusions from Thaler (1981)

Conclusion #1: The amount matters --- there is more discounting for smaller amounts ("magnitude effect" ). 

--

But the __key__ conclusion is: 

Conclusion #2: For either amount, discount rates are higher in the short run than in the long run (sometimes referred to as "declining discount rates" ).

--

This finding is inconsistent with exponential discounting!

---
class: MSU
# Some "Preference Reversals"

Consider the following (hypothetical) choice scenarios: 

__Choice 1:__

$$\left[ 10\text{ M&Ms now}\right] \qquad \text{vs.} \qquad  \left[ 15\text{
M&Ms tomorrow}\right]$$

--

__Choice 2:__
$$\left[ 10\text{ M&Ms in 7 days}\right] \qquad \text{vs.} \qquad  \left[ 15\text{
M&M in 8 days}\right]$$

--

A plausible pattern:

$$(10 \text{ M&Ms now } ) \succ (15 \text{ M&Ms tomorrow})$$

$$( 15 \text{ M&Ms in eight days} ) \succ (10 \text{ M&Ms in seven days})$$
---
class: MSU 
# Using these "Preference Reversals"

Implication: Under the usual assumptions, this pattern implies 

$$D(0)10 > D(1)15\qquad \Longleftrightarrow \qquad \frac{D(0)}{D(1)}>1.5$$
--

$$ D(7)10 < D(8)15\qquad \Longleftrightarrow \qquad \frac{D(7)}{D(8)}<1.5$$ 
--

$$ \text{Hence}\text{: } \frac{D(0)}{D(1)}>\frac{D(7)}{D(8)} $$ 
--

**Conclusion:** If you exhibit this pattern, then you are more impatient toward the now-vs.-near-future tradeoff than you are toward the near-future-vs.-further-future tradeoff.

---
class: MSU 
# Real Evidence of "Preference Reversals"

In fact, Kirby & Herrnstein (1995) show that, for 23 of their 24 subjects, they can make the subject exhibit this type of "preference reversal" . 

--

For instance, for each of these 23 subjects, they find an $x>0$ and a $y>0$
such that the subject's preferences are 

$$( \text{45 now }) \succ ( \text{52 in } y \text{ days})$$  


$$( \text{ 45 in } x \text{ days} ) \succ ( \text{ 52 in } y \text{ days})$$  
--

Much as above, under the usual assumptions, this implies


$$\frac{D(0)}{D(y)}>\frac{D(x)}{D(x+y)}$$

--

Again, more impatient toward the now-vs.-near-future tradeoff than toward the near-future-vs.-further-future tradeoff.

---
class: MSU 
# Evidence of "Hyperbolic Discounting"

Another approach to the same type of data is to directly compare two functional forms: 

1.  Exponential Discounting: $D(x)=e^{-kx}$ 

2. Hyperbolic Discounting: $D(x)=\frac{1}{1+kx}$ 

--

In these comparisons, the answer is that hyperbolic discounting is virtually always a better fit (occasionally, they're equally good).

---
class: MSU 
# Evidence of "Hyperbolic Discounting"
\frametitle{Evidence of " Hyperbolic
Discounting" }

For instance, Kirby (1997) elicited WTP's for $20 to be received in $x$ days, where each subject answered for every odd $x$ between 1 and 29. 

--

He then tested for each subject whether their discount function was better fit by the exponential functional form or the hyperbolic functional form. 

--

**Results:** Hyperbolic was a better fit for 59 of 67 subjects, exponential was a better fit for 6 subjects, and for 2 subjects the functions were equally good.

---
class: MSU 
# Conclusions from the Evidence

In the experimental data, there seems to be a key feature that virtually always holds: 

- Discount rates are higher in the short run than in the long run (sometimes referred to as "declining discount
rates" ). 

--

In terms of our notation, the evidence seems to suggest:

$$\frac{D(0)}{D(1)}>\frac{D(1)}{D(2)}>\frac{D(2)}{D(3)}>...$$

--

Note: This suggests a time inconsistency of an impulsive nature: 

--

- When thinking about some future period, you'd like to behave relatively patiently; but when the time comes to actually choose your behavior, you want to behave relatively impatiently.
