---
title: "Choice over Time: Introduction"
subtitle: "EC404; Spring 2024"
author: "Prof. Ben Bushong"
date: "Last updated `r format(Sys.Date(), '%B %d, %Y')`"
output:
  xaringan::moon_reader:
    lib_dir: libs
    yolo: false
    css: [default, metropolis, "EC404.css"]
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
    incremental: true
name: "Cover"


---
```{r setup, include=FALSE}

options(htmltools.dir.version = FALSE)
options("getSymbols.warning4.0"=FALSE)


library(here) # creates absolute paths, but those won't work in html
library(knitr)
library(kableExtra)
library(tidyverse)
library(wooldridge)
require(lubridate)
require(scales)
require(broom)
require(visualize)

require(wbstats)
require(lmtest)
require(sandwich)
require(car)
require(quantmod)
require(patchwork)



# https://yihui.org/knitr/options/
opts_chunk$set(
  fig.align="center",
  #dpi=300,
  fig.path='figs/', # where figs are rendered
  cache=F,
  echo=F,
  message = F,
  warning = F
  )

```

layout: true

<div class="msu-header"></div>
<div style = "position:fixed; visibility: hidden">
$$\require{color}\definecolor{yellow}{rgb}{1, 0.8, 0.16078431372549}$$
$$\require{color}\definecolor{orange}{rgb}{0.96078431372549, 0.525490196078431, 0.203921568627451}$$
$$\require{color}\definecolor{MSUgreen}{rgb}{0.0784313725490196, 0.52156862745098, 0.231372549019608}$$
</div>

<script type="text/x-mathjax-config">
MathJax.Hub.Config({
  TeX: {
    Macros: {
      yellow: ["{\\color{yellow}{#1}}", 1],
      orange: ["{\\color{orange}{#1}}", 1],
      MSUgreen: ["{\\color{MSUgreen}{#1}}", 1]
    },
    loader: {load: ['[tex]/color']},
    tex: {packages: {'[+]': ['color']}}
  }
});
</script>

<style>
.yellow {color: #FFCC29;}
.orange {color: #F58634;}
.MSUGreen {color: #14853B;}
</style>


```{r flair_color, echo=FALSE}
library(flair)
yellow <- "#FFCC29"
orange <- "#F58634"
MSUGreen <- "#14853B"
```


```{r, eval=FALSE}

#BELOW IS WHERE YOU ADD CONTENT. KEEP ABOVE CONSTANT EXCEPT TITLE.

```

---
class: MSU
# Overview of Topic 2

## Topic 2: 
### Intertemporal Choice or Making Choices Over Time

Many interesting questions in economics involve choice over time:

- How do people allocate their wealth between current consumption and future consumption?

- How do people decide when to work on tasks?

- For goods that yield short-term consumption utility but generate negative consequences in the long-term---e.g., alcohol, cigarettes, potato chips---how do people trade off the short-term benefits vs. the long-term costs?



---
class: MSU
# Overview of Topic 2


The standard model ("exponential discounting") assumes:

1. People treat time in a relatively even-handed manner.

2. People carry out their plans.

3. People know what they'll like in the future.

---
class: MSU
# Today's Lecture

### Warm-Up: Interest Rates, Compounding, PDV

Let's understand some of the precursors to the standard model and (quickly) do some example problems. (This is a great time to ask questions.)

### The Standard Model (Today + Thursday)

---
class: MSU
# Interest Rates and Compounding

**Example A:** Suppose you put $1000 into a bank account that pays 10% interest per year.

- After 1 year, you'll have $1000 * (1.10) = $1100 .

- After 2 years, you'll have $1100 * (1.10)= $1210 .

- After 3 years, you'll have $1210 * (1.10)= $1331 .

--

More generally:

If you put $P$ into a bank account that pays interest rate $r$ per year, its future value in $T$ years will be $P\ast(1+r)^{T}$ .

---
class: MSU
# Interest Rates and Compounding

### Definitions (easy; hopefully not new)

*Compound interest* is interest paid on past interest earned.

--

*Compounding* is earning interest on past interest earned.

--

The **frequency of compounding** is the frequency at which interest is credited to your account (after which it's starts earning compound interest).

--

Our example above implicitly assumed yearly compounding. Of course, we could have more frequent compounding....

---
class: MSU
# Interest Rates and Compounding

**Example B:** Suppose you put $1000 into a bank account that pays a 10% annual interest rate that is compounded every six months.

Because a 10% annual interest rate implies a 5% semi-annual interest rate:

- After 6 months, you'll have $1000 \ast (1.05)= 1050$ .

- After 1 year, you'll have $1050 \ast (1.05)= 1102.50$ .

--

**Example C:** Suppose you put $1000 into a bank account that pays a 10%
annual interest rate that is compounded every month.

Because a 10% annual interest rate implies a $0.8\bar{3}$ % monthly interest rate:

- After 1 year, you'll have $(1000)\ast(1.008\bar{3})^{12}=1104.71$ .



---
class: MSU
# Interest Rates and Compounding

More generally, if you put $P$ into a bank account that pays an annual interest rate of $r$ that is compounded $n$ times per year:

- Its future value after 1 year will be $(P)\ast (1+r/n)^{n}$ .

- Its future value after $T$ years will be $(P)\ast \left[ (1+r/n)^{n} \right] ^{T}$ .

--

- Note: For continuous compounding, $\lim_{n\rightarrow \infty}(1+r/n)^{n}=e^{r}$ and $\lim_{n\rightarrow \infty }\left[ (1+r/n)^{n}\right] ^{T}=e^{rT}$ .

---
class: MSU
# Discrete-Time Models

Suppose there is some set of periods $0,1,2,...,T$ (perhaps $T=\infty$ ).

- Note: The length of a period might be one year, one month, one day, or whatever is most appropriate for the particular application.

Suppose there is a per-period interest rate $r$, and interest is compounded every period.

If $P_{t}$ is the principal in your bank account in period $t$, then:


- $P_{1}=(1+r)\ast P_{0}$

- $P_{2}=(1+r)^{2}\ast P_{0}$

- $P_{t}=(1+r)^{t}\ast P_{0}$

- $P_{6}=(1+r)\ast P_{5}$

- $P_{6}=(1+r)^{4}\ast P_{2}$

- $P_{t+x}=(1+r)^{x}\ast P_{t}$

---
class: MSU
# Present Discounted Value (PDV)

Suppose that you will be paid $1100 one year from today. If the market interest rate is 10% (and yearly compounding), how much is this future payment be worth to you now?

--

We can answer this question by asking how much you could borrow now such that you would have to pay back exactly $1100 in one year.

- Answer: $1000 --- because $(1.10)\ast(1000)=1100$ .


--

**Definition**: Given per-period interest rate $r$, the *present discounted value* (or sometimes just *present value* or *PDV*) of $P$ to be paid $T$ periods in the future is

$$\frac{P}{(1+r)^{T}}$$


---
class: MSU
# Present Discounted Value (PDV)

Some $PDV$ 's for $P=1000$ and yearly compounding:


| $r$ | 1 Year | 2 Years | 3 Years | 10 Years | 20 Years |
|:---:|:------:|:-------:|:-------:|:--------:|:--------:|
| 3% | $971 | $943 | $915 | $744 | $554 |
| 4% | $962 | $925 | $889 | $676 | $456 |
| 5% | $952 | $907 | $864 | $614 | $377 |
| 6% | $943 | $890 | $840 | $558 | $312 |
| 7% | $935 | $873 | $816 | $508 | $258 |


---
class: MSU
# PDV of a Stream of Payoffs

Suppose that you will be paid $1100 one year from today, another $1100 two years from today, and yet another $1100 three years from today.

--

If the market interest rate is 10% (and yearly compounding), how much is this stream of payoffs worth to you now?

**Answer:** Add up the individual $PDV$ bit-by-bit:

$$PDV=\frac{\$1100}{(1.10)}+\frac{\$1100}{(1.10)^{2}}+\frac{\$1100}{(1.10)^{3}}=\$2735.54.$$

--

More generally: Given per-period interest rate $r$, a stream of future revenues $(R_{1},R_{2},...,R_{N})$ (where revenue $R_{n}$ is received in period $n$) has a present discounted value of:


$$PDV=\frac{R_{1}}{(1+r)}+\frac{R_{2}}{(1+r)^{2}}+...+\frac{R_{N}}{(1+r)^{N}}.$$

---
class: MSU
# Time-Varying Interest Rates

**Definition:** The *period-* $t$ *interest rate* $r_{t}$ is the interest rate between period $t$ and period $t+1$. In other words, if in period $t$ your principal is $P_{t}$, then in period $t+1$ it becomes $P_{t+1}=(1+r_{t})P_{t}$.

--

Hence, if $P_{t}$ is the principal in your bank account in period $t$, and if your bank account pays per-period interest rates $(r_{t},r_{t+1},...)$,
then:


- $P_{t+1}=(1+r_{t})P_{t}$.

- $P_{t+2}=(1+r_{t+1})P_{t+1}=(1+r_{t+1})(1+r_{t})P_{t}$.

- $P_{t+3}=(1+r_{t+2})P_{t+2}=(1+r_{t+2})(1+r_{t+1})(1+r_{t})P_{t}$.

- And so on....

---
class: MSU
# Time-Varying Interest Rates

Given per-period interest rates $(r_{t},r_{t+1},...)$, a stream of future
revenues $(R_{t+1},R_{t+2},R_{t+3})$ has a present discounted value of

$$PDV = \frac{R_{t+1}}{(1+r_{t})}+\frac{R_{t+2}}{(1+r_{t})(1+r_{t+1})}+\frac{R_{t+3}}{(1+r_{t})(1+r_{t+1})(1+r_{t+2})}.$$

---
class: inverseMSU 
# End of Warm-Up 

As always, you will *not* need to memorize any of these equations. But we want to remind ourselves how to think about choices across time. 

- How do you assess today versus tomorrow? 
- What is the "correct" weight to put on money today versus money tomorrow? 
- What determines this tradeoff? 

We'll explore these questions (and many more) coming up. 

