---
title: Syllabus
slug: syllabus
publishDate: "2025-01-14"
citeproc: true
output:
  blogdown::html_page:
    toc: true
    fig_caption: false
---

```{r setup, message=FALSE, warning=FALSE, include=FALSE}
suppressPackageStartupMessages(library(dplyr))
library(tibble)
library(pander)
```

`r blogdown::shortcode("courseinfo")`

### Office Hours

Office hours for this course are by request and integrated into the course. 

## What Is This Course and Can / Should You Take It?

This course will introduce students to decades of psychological research indicating systematic departures from classical economic assumptions. Furthermore, this course will demonstrate how that research can be translated into formal models that can be incorporated into economics. Topics include ways utility theory can be improved—such as incorporating reference dependence, social preferences, self image, and other belief-based tastes—and ways we can relax assumptions of perfect rationality—such as incorporating focusing effects, limited attention, biased prediction of future tastes, present-biased preferences, and biases in probabilistic judgment. The course will emphasize (a) careful interpretation of new evidence, (b) formalizing this evidence into models that can, with discipline and rigor, generate sharp predictions using traditional economic approaches, and (c) exploring economic implications of those models presented. Although the course will primarily emphasize (b), it is intended to be useful to students whose interests lie anywhere in this spectrum.

The course is intended for students in the Economics program and others who have a solid background in microeconomic theory at the level of intermediate courses in these programs. This course contains material and approaches targeted at those in the economics major.  In this course, you will be required to use a fair amount of algebra and some elementary calculus, which demands that you can take derivatives and derive an optimum.

> Enrollment for credit in this course is simply not suitable for those unprepared in or uninterested in economic theory no matter the intensity of interest in psychology or behavioral economics.  Really.

The course focuses on a mix of formal modeling and a careful review of existing evidence.  The course will involve more words and more intuition and more classical economic theory than the typical topics course.

The topics covered in this course are listed later in the syllabus. There is no textbook for the course. When available, I will give links to papers that contain the formal models, but I will not require that you read those papers. The important thing to remember is that, since there is no book for this course, you will rely on your course notes much more than you might in other classes. I cannot emphasize this enough: learning from course materials is extremely hard if you are not present during the initial presentation of those materials. Thus, you are much more likely to succeed if you attend class.

If—after you have read the syllabus—you have any questions about whether this course is appropriate for you, please come talk to me.  Guests are always permitted to attend the lectures (I may ask you to introduce them).

## Non-Detailed Course Psuedo-Description

This course is designed as an introduction to modeling psychological phenomena that are not yet totally integrated into mainstream economic analysis. Thus the material in this course is not entirely like what you’ve seen in most of your other economics courses.  But it is not an alternative to mainstream economics.  It is only about improving the psychological realism of formal economics, and we will use classical economic approaches to improve our answers to classical economic questions. Most topics will be introduced by first going through the classical model, then highlighting empirical findings that do not fit with that model, and finally exploring alternative approaches. Like all other courses, this course does not cover all topics that might be of interest, and we will not spend time exploring methodological issues. However, as undergraduates studying the field of economics, methodology can and should be your primary focus. Economic methodology will provide the backbone for your future careers and, should you choose to pursue it, any graduate work in economics.

__This course is__

  - not about non-psychological models of bounded rationality

We won’t consider models of bounded rationality (based on computer science, artificial intelligence, etc.) that are meant to capture cognitive limits of economic actors, but not based on evidence that humans think this way.  In some arenas I think it makes tremendous sense to focus on these alternative models of bounded rationality, and more generally this can be a very useful research agenda.  But that’s not what this course is about.  We will consider those models based on research inspired by the empirical evidence of what humans are like.

  - not about experimental economics as such

We will examine the experimental evidence of the topics covered in this course.  But the course won’t be about experimental methods per se. This course is meant to use the results from experiments to motivate new economic assumptions, and to emphasize the potential for non-experimental research in these topics.  We won’t study experiments testing economic institutions in the laboratory, except insofar as they are either motivated by or informative about the underlying psychology of economic actors.

  - absolutely not an alternative to mainstream economics

In the most important senses, the course won’t at all be a departure from mainstream economics. I believe that the best way for economists to do economics in general, and the best way for us to use this material in particular, is with careful formal theory and (albeit absent from this course) statistical analysis.  In these regards, the course will be purposely, pointedly, persistently, proudly, and ponderously mainstream.


## Contacting Me
I have moved this up to an odd spot in a (likely unsuccessful) attempt to minimize our collective headache.

Email is a blessing and a curse. Instant communication is wonderful, but often email is the wrong medium to have a productive conversation about course material. Moreover, I get a **lot** of emails. This means that I am frequently triaging emails into two piles: "my house is burning down" and "everything else". Your email is unlikely to make the former pile. So, asking questions about course material is always best done in-class or in office hours. Students always roll their eyes when professors say things like that, but it’s true that if you have a question, it’s very likely someone else has the same question.

That said, it benefits us both if any emails you send are clear and effective. There’s an (unfunny) joke in academia that professors (i) read an email until they find a question; (ii) respond to that question and; (iii) ignore the rest of the email. I won’t do this, but I think it is helpful to assume that the person on the receiving end of an email will operate this way.

Some general tips:

- **Always** include ``[EC404]`` in your subject line (brackets included).
- Use a short but informative subject line. For example: ``[EC404] Midterm Exam Grading``
- Use your University-supplied email for University business. This helps me know who you are.
- One topic, one email. If you have multiple things to discuss, and you anticipate followup replies, it is best to split them into two emails so that the threads do not get cluttered.
- Ask direct questions. If you’re asking multiple questions in one email, use a bulleted list.
- Don’t ask questions that are answered by reading the syllabus. This drives me nuts.
- I’ve also found that students are overly polite in emails. I suppose it may be intimidating to email a professor, and you should try to match the style that the professor prefers, but I view email for a course as a casual form of communication. Said another way: get to the point. Students often send an entire paragraph introducing themselves, but if you use your University email address, and add the course name in the subject, I will already know who you are. Here’s an example of an excellent email:

> Subject: [EC404] Assignment 2 Typo
>
> Hi Prof. Bushong,
>
> There seems to be a typo in the assignment on Question 2. The problem says to take a derivative with respect to a variable that doesn't exist. Can you correct this or specify what you mean?
>
> Thanks,
> Student McStudentFace

Now on to your regularly scheduled syllabus.

## Success in This Course

I *promise*, you are equipped to succeed in this course.

That said, learning sucks.

True learning definitionally means you didn't know something. When it comes to psychological factors, people are often hesitant to say they don't know something. (Or will insist they do in fact know something when they really don't.)

This can be exacerbated in this class by the intersection of tough psychology and challenging mathematics. In my experience, many students are not proficient with high-school algebra. This isn't an insult: it means your teachers sucked.

> It's easy when you start out in this class to get really frustrated and think, "Oh it's me, I'm really stupid," or, "I'm just not good at math." But, that is absolutely not the case. Everyone gets frustrated. It's just a natural part of learning. Your brain doesn't want to work: it's lazy. So, forcing it to exert effort is stressful and, at times, painful. The pain happens to everyone (even the best students) and gets less and less over time. Don't blame yourself. Just take a break, do something fun, and then come back and try again later.

### Problem Sets

The primary mode of learning in this course is through (difficult) problem sets.

There will be six problem sets.  Problems will range in difficulty from moderately hard to impossibly, annoyingly hard, so don’t panic if you struggle with them. You are encouraged to work together on the problem sets, but you are not allowed to read others’ answers from past or present problem sets, and should hand in solutions separately that reflect your own understanding.  You should acknowledge collaborators and describe the extent of collaboration at the top of the relevant problem sets. While collaboration is allowed, directly copying someone else's work is not.

Problem sets are due at the beginning of lecture the day they are due. Late problem sets will not be accepted.  Please staple your problem set together (paper clips and folded corners are not staples). If you cannot attend lecture the day a problem set is due, you can turn it in to either my office or the TA’s office.

There is a limited amount of time we can spend grading the problem sets. Problems will frequently require substantial math; you are welcome to hand in all the work you did to reach an answer, but that is not required. Clearly indicate your answer by labeling (if typed) or circling (if handwritten). You will not receive feedback on your problem sets, but you will receive an answer sheet.

Problem sets will be graded using a simple system. Problem sets that are late will be given a 0. We will grade problem sets in the most forgiving way possible, but you must complete each problem. That said, lengthy explanations of how you arrived at a wrong answer will not necessarily entitle you to full credit, regardless of the length. A good-faith effort will generally earn you an 10. Exceptional work (with mostly or entirely correct answers) will earn 11.5.

In the end, problem sets aren’t a big component of your grade. Their primary purpose is to teach you the actual material and to prepare you for exams---hence the relaxed grading and feedback. There will be two midterms and a final exam (see next page). These will draw heavily from the problem sets, and therefore a deep understanding of the problem sets is the best way to succeed in this course. 

### Evaluations and Grades

Your grade in this course will be based on attendance + participation, assignments, three exams, and a writing project. The general breakdown for grades will be 10% for attendance, 30% for assignments, 50% for exams, and 10% for a "paper". Details on the paper will be distributed midway through the course – it is designed not to deter anyone from taking this course who might otherwise be interested.


<!-- Each exam will have a variable grading weight. -->

<!-- - Midterm 1:	5-17.5%  -->
<!-- - Midterm 2:	5-17.5% -->
<!-- - Final Exam:	14-39% -->

<!-- To calculate your overall exam score, I will automatically choose the grading weights most advantageous to you. For example, if the final exam is your best score, then your grading weights will be 35% for the final exam and 5% for each midterm.  If the final exam is your worst score, then your grading weights will be 10% for the final exam and 17.5% for each prelim. This scheme will help mitigate the effects of one-time disasters, including poor performance early in the course (it happens). -->

**I reserve the right to give higher grades**, but the above is guaranteed. Any adjustments will be done consistently, with course attendance and completion percentage (e.g., did you do all the assigned work?) serving as tiebreakers where needed.

Attendance and participation grades will be based on completion of in-class assignments. This is intended to be an easy way to get points.

More in-depth descriptions for all the assignments will be on the [assignments page](/assignment/). As the course progresses, the assignments themselves will be posted to the appropriate place.

**To Recap:**

```{r assignments-grades, include=FALSE}
assignments <- tribble(
  ~Points,  ~Assignment,
  20      , "Class Participation",
  20*3    , "Assignments",
  100     , "Exams",
  20      , "Paper",
) %>%
  mutate(Percent = Points / sum(Points),
         Percent = scales::percent(Percent, accuracy = 1))

grading <- tribble(
  ~Grade, ~Range,  ~Grade1, ~Range1,
  "4.0",  "92-100%", "2.0",  "72-76%",
  "3.5", "87-91%",  "1.5", "67-72%",
  "3.0", "82-87%",  "1.0", "62-67%",
  "2.5",  "77-81%",  "0.0",  "bad-66%",
)
```

<div class="centered-table">

```{r show-assignments-table, echo=FALSE, results="asis"}
assignments_total <- tibble(Assignment = "Total",
                            Points = sum(assignments$Points),
                            Percent = NA)

bind_rows(assignments, assignments_total) %>%
  dplyr::select(Assignment, Points, Percent) %>%
  pandoc.table(justify = "lcc", missing = "—")
```
</div>

<div class="centered-table">

```{r show-grades-table, echo=FALSE, results="asis"}
grading %>%
  pandoc.table(justify = "lclc",
               col.names = c("Grade", "Range", "Grade", "Range"))
```

</div>

All grades are considered final. Any request for a re-grade beyond simple point-tallying mistakes will require that the entire assignment be re-graded. Any points previously awarded may be changed in either direction in the re-grade

### Academic Honesty

Violation of [MSU's Spartan Code of Honor](http://asmsu.msu.edu/home/initiatives/spartan-code-of-honor/) will result in a grade of 0.0 in the course. Moreover, I am required by MSU policy to report suspected cases of academic dishonesty for possible disciplinary action. Concretely: you may **not** use old course materials, even if you find them online or a friend gives them to you. I am good at changing things *just slightly* so as to catch those whose work is not their own.^[So just don't cheat or plagiarize. This is an easy problem to avoid.]

## Resources

Mental health issues or stressful events may lead to diminished academic performance or otherwise reduce your ability to participate in daily activities. Services are available to assist you with addressing these and other concerns you may be experiencing. You can learn more about the broad range of confidential mental health services available on campus via the Counseling & Psychiatric Services (CAPS) website linked [here](www.caps.msu.edu).

### Accommodations

This class is designed to be fairly accommodating without a student asking. However, if you need a special accommodation for a disability, religious observance, or have any other concerns about your ability to perform well in this course, please contact me immediately so that we can discuss the issue and make appropriate arrangements.

Michigan State University is committed to providing equal opportunity for participation in all programs, services and activities. Requests for accommodations by persons with disabilities may be made by contacting the Resource Center for Persons with Disabilities at 517-884-RCPD or on the web [(here)](rcpd.msu.edu). Once your eligibility for an accommodation has been determined, you will be issued form that specifies any accomodations that you require. Please present this form to me at the start of the term and/or two weeks prior to the accommodation date (test, project, etc). Requests received after this date will be honored whenever possible but presenting the form early is the best way to ensure that your needs are met.


### A Note on Things Right Now

Things for you might be especially hard right now. I get that.

**I'm fully committed to making sure that you learn everything you were hoping to learn from this class.** I will make whatever accommodations I can to help you finish your exercises, do well on your projects, and learn and understand the class material. Under ordinary conditions, I am flexible and lenient with grading and course expectations when students face difficult challenges. Given the challenges of the past two years, that flexibility and leniency is intensified.

If you feel like you're behind or not understanding everything, **do not suffer in silence.** *Please* contact me. I'm available via [e-mail](mailto:bbushong@msu.edu); or come see me.

As a final note: if you've been on this planet for the past few years, you're well aware that we've had some wild and wacky changes that we had to confront. If such a change happens, I will do my best to ensure that your experience in this course does not suffer. That said, I hope you will extend a bit of leeway in such a circumstance.

### Mandatory Reporting

Materials submitted for this class are generally considered confidential pursuant to the University's student record policies.  However, students should be aware that University employees, including instructors, may not be able to maintain confidentiality when it conflicts with their responsibility to report certain issues to protect the health and safety of MSU community members and others.  As the instructor, I must report the following information to other University offices (including the Department of Police and Public Safety) if you share it with me:
- Suspected child abuse/neglect, even if this maltreatment happened when you were a child;
- Allegations of sexual assault, relationship violence, stalking, or sexual harassment; and
- Credible threats of harm to oneself or to others.

These reports may trigger contact from a campus official who will want to talk with you about the incident that you have shared.  In almost all cases, it will be your decision whether you wish to speak with that individual.  If you would like to talk about these events in a more confidential setting, you are encouraged to make an appointment with the MSU Counseling and Psychiatric Services.

## Miscellanea

D2L will be used sparingly---primarily for submission of assignments and distribution of grades. This course website is the primary source of information.

### Using Office Hours

Please use my office hours. It would be remarkable if you didn’t need some assistance with the material, and I am here to help.  One of the benefits of open office hours is to accommodate many students at once; if fellow students are "in my office", please join in and feel very free to show up in groups. Office hours will move around a little bit throughout the semester to attempt to meet the needs of all students. 

In addition to drop-in office hours, I always have sign-up office hours for advising and other purposes. As a general rule, please first seek course-related help from the drop-in office hours. However, if my scheduled office hours are never feasible for you, let me know, and then I may encourage you to make appointments with me. I ask that you schedule your studying so that you are prepared to ask questions during office hours – office hours are not a lecture and if you’re not prepared with questions we will end up awkwardly staring at each other for an hour until you leave.

Some gentle requests regarding office hours and on contacting me. First, my office hours end sharply at the end, so don’t arrive 10 minutes before the scheduled end and expect a full session. Please arrive early if you have lengthy questions, or if you don’t want to risk not having time due to others’ questions. You are free to ask me some stuff by e-mail, (e.g. a typo or something on a handout), but please know e-mail sucks for answering many types of questions. “How do I do this assignment?” or “What’s up with calculus?” are short questions with long answers. Come to office hours. 


### Letters of Recommendation / References
I'm not doing letters of recommendation at the moment. It's not you, it's me. If you want to convince me otherwise, you may try, but... I'm not optimistic. 
